Making Higher-Order Superposition Work

### 2022-03-24 scratch/paper.tex
more
prepare ehoh2

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required

%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
\documentclass[smallcondensed,draft]{svjour3}     % onecolumn (ditto)
%\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage[utf8]{inputenc}
\usepackage[misc]{ifsym}
\usepackage[final]{graphicx}
\usepackage{array}
\usepackage{multirow}
\usepackage{cite}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage{prftree}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{enumerate}
\usepackage{stmaryrd}
\usepackage{booktabs}
\usepackage{algorithmicx}
\usepackage{algorithm}% http://ctan.org/pkg/algorithms
\usepackage[noend]{algpseudocode}% http://ctan.org/pkg/algorithmicx
\usepackage[
   a4paper,
   pdftex,
   pdftitle={Making Higher-Order Superposition Work},
   pdfauthor={Petar Vukmirovi\'c, Alexander Bentkamp, Jasmin Blanchette, Simon Cruanes, Visa Nummelin, and Sophie Tourret},
   pdfkeywords={},
   pdfborder={0 0 0},
   draft=false,
   bookmarksnumbered,
   bookmarks,
   bookmarksdepth=2,
   bookmarksopenlevel=2,
   bookmarksopen]{hyperref}

\usepackage{mathptmx}      % use Times fonts if available on your TeX system

\DeclareSymbolFont{letters}{OML}{txmi}{m}{it} %% for Greek

% insert here the call for the packages your document requires
%\usepackage{latexsym}
% etc.
%
% please place your own definitions here and don't use \def but
% \newcommand{}{}
%
% Insert the name of "your journal" with
\journalname{Journal of Automated Reasoning}

\makeatletter

\def\vthinspace{\kern+0.083333em}
\newcommand\CORR{{(\vthinspace\Letter\vthinspace)}}

\def\orcid#1{{\href{http://orcid.org/#1}{\protect\raisebox{-1.25pt}{\protect\includegraphics{orcid.pdf}}}}}

@@

%%% HACK Springer's style to get bold subsection headings --JB
\def\subsection{\@startsection{subsection}{2}{\z@}%
    {-21dd plus-8pt minus-4pt}{10.5dd}
     {\normalsize\bfseries\boldmath}}

\spnewtheorem{definition}[theorem]{Definition}{\bfseries}{\rmfamily}
\spnewtheorem{exa}[theorem]{Example}{\bfseries}{\rmfamily}

@@

\title{Making Higher-Order Superposition Work}
% \subtitle{Do you have a subtitle?\\ If so, write it here}

%\titlerunning{Short form of title}        % if too long for running head

\author{Petar~Vukmirović~\orcid{0000-0001-7049-6847}\and
Alexander~Bentkamp~\orcid{0000-0002-7158-3595}\and
Jasmin~Blanchette~\orcid{0000-0002-8367-0936}\and
Simon~Cruanes~\orcid{0000-0003-3969-5850}\and
Visa~Nummelin~\orcid{0000-0003-0078-790X} \and
Sophie~Tourret~\orcid{0000-0002-6070-796X}}

\authorrunning{Petar Vukmirović et al.} % if too long for running head

\institute{
  Petar~Vukmirovi\'c \CORR \and
  Alexander Bentkamp\and
  Jasmin Blanchette\and
  Visa Nummelin\at%
    Vrije Universiteit Amsterdam, Amsterdam, the Netherlands\\
    \email{\{p.vukmirovic,a.bentkamp,j.c.blanchette,visa.nummelin\}@vu.nl}
  \and
  Jasmin Blanchette\and Sophie Tourret \at%
    Universit\'e de Lorraine, CNRS, Inria, LORIA, Nancy, France
    \\
    \email{\{jasmin.blanchette,sophie.tourret\}@inria.fr}
  \and
  Jasmin Blanchette\and Sophie Tourret \at  Max-Planck-Institut f\"ur Informatik, Saarbr\"ucken,~Germany \\
  \email{\{jblanche,stourret\}@mpi-inf.mpg.de}
  \and
  Simon Cruanes \at Imandra, Austin, Texas, USA \\
  \email{simon@imandra.ai}
}

\date{Received: date / Accepted: date}
% The correct dates will be entered by the editor

@@

\begin{abstract}
Superposition is among the most successful calculi for first-order logic. Its
extension to higher-order logic introduces new challenges such as infinitely
branching inference rules, new possibilities such as reasoning about
Booleans, and the need to curb the explosion of specific higher-order rules. We
describe techniques that address these issues and extensively evaluate their
implementation in the Zipperposition theorem prover. Largely thanks to their use,
Zipperposition won the higher-order division of the CASC-J10 competition.
\end{abstract}

\newcommand{\ifalse}{\pmb\bot}
\newcommand{\itrue}{\pmb\top}
\newcommand{\inot}{\pmb{\neg\,}}
\newcommand{\iand}{\pmb\land}
\newcommand{\ior}{\pmb\lor}
\newcommand{\iimplies}{\pmb\rightarrow}
\newcommand{\iequiv}{\pmb\leftrightarrow}
\newcommand{\iforall}{\pmb\forall}
\newcommand{\iexists}{\pmb\exists}
\newcommand{\ieq}{\pmb\approx}
\newcommand{\ineq}{\pmb{\not\approx}}

\newcommand{\comb}[1]{\ensuremath{\mathsf{#1}}}
\newcommand{\neglit}[1]{\ensuremath{#1 \eq \ifalse}}
\newcommand{\poslit}[1]{\ensuremath{#1 \eq \itrue}}
\newcommand{\skbci}{\textsf{SKBCI}}
\def\negvthinspace{\kern-0.083333em}

\newcommand{\cst}[1]{\ensuremath{\mathsf{#1}}}
\newcommand{\var}[1]{{\mathit{#1}}}
\newcommand{\typ}[1]{{\mathit{#1}}}
\newcommand\foralltynospace[1]{\mathsf{\Pi}#1.}
\newcommand\forallty[1]{\foralltynospace{#1}\;}
\newcommand{\typeargs}[1]{{\negvthinspace\langle#1\rangle\negvthinspace}}
\let\oldDelta=\Delta
\let\oldSigma=\Sigma
\renewcommand\Sigma{\mathrm{\oldSigma}}
\newcommand{\Sigmaty}{\Sigma_\mathsf{ty}}
\newcommand{\VV}{\mathscr{V}}
\newcommand{\Vty}{\VV_\mathsf{ty}}

% Equality predicate
\newcommand{\eq}{\approx}
\newcommand{\noteq}{\not\eq}
\newcommand{\lland}{\mathrel\land}
\newcommand{\llor}{\mathrel\lor}
\newcommand{\ccup}{\mathrel\cup}
\newcommand{\ccap}{\mathrel\cap}

\newcommand{\tuple}[2]{\overline{#1}_{#2}}
\newcommand{\tuplen}[1]{\overline{#1}_{n}}

% Names of inference rules
\newcommand{\infname}[1]{\textsc{#1}}

% Inference rule
\newcommand{\namedinference}[3]{\prftree[r]{\relax{\infname{#1}}}{\strut#2}{\strut#3}}
\newcommand{\inference}[2]{\namedinference{}{\strut#1}{\strut#2}}

% Simplification rule
\newcommand{\namedsimp}[3]{\prftree[d][r]{\relax{\infname{#1}}}{\strut#2}{\strut#3}}
\newcommand{\simp}[2]{\namedinference{}{\strut#1}{\strut#2}}
\newcommand{\lsupraw}{$\lambda$-super\-po\-si\-tion}
\newcommand{\lsup}{{\color{black}\lsupraw}}
\newcommand{\lsup}{{\color{black}Boolean-free \lsupraw}}
\newcommand{\osup}{{\color{black}Boolean \lsupraw}}

\newcommand\NumberOK[1]{#1}
\newcommand\NumberNOK[1]{\colorbox{red}{#1}}
\newcommand{\ourpara}[1]{\paragraph{\upshape\bfseries#1}}

\hyphenation{su-per-po-si-tion Zip-per-posi-tion tab-leau tab-leaux Bent-kamp ab-strac-tion
  Bool-ean}

\newcommand{\confrep}[2]{#2}

\section{Introduction}
\label{sec:satfol:ho-tech:intro}

%In recent decades,
Superposition-based first-order automatic theorem provers
have emerged as useful reasoning tools. They dominate at the annual CASC
\cite{gs-2016-casc} theorem prover competition, having always won the
first-order theorem division. They are also used as backends to proof assistants
\cite{ck-18-coqhammer,ku-15-holyhammer,pb-12-sh}, automatic
higher-order theorem provers \cite{sb-21-leo3}, and software verifiers
\cite{fp-13-why3}.

The superposition calculus has only recently been extended
to higher-order logic (more precisely, extensional simple type theory
\cite{henkin-1950-completeness}), resulting in
\emph{\lsup} \cite{bentkamp-et-al-2021-lamsup-journal}, which we developed
together with Waldmann, as well as \emph{combinatory superposition}
\cite{br-20-full-sup-w-combs} by Bhayat and Reger. Although these two
calculi do not support an interpreted Boolean type,
they can be extended by ad hoc rules \cite{our-bool-paper} that support
most of the Boolean reasoning necessary in practice.

Both higher-order superposition calculi were designed to gracefully
extend first-order reasoning. As most steps in higher-order
proofs tend to be essentially first-order, extending the most successful first-order
calculus to higher-order logic seemed worth trying.
Our first attempt at testing this idea was in 2019:
Zipperposition~1.5, based on \lsup, finished third
in the higher-order theorem division of CASC-27 \cite{gs-19-casc27},
12~percentage points behind the winner, the tableau prover Satallax 3.4 \cite{cb-2013-satallax}.

Studying the competition results, we found that higher-order tableaux have
some advantages over higher-order superposition. To bridge the gap, we developed
techniques and heuristics that simulate tableaux in the
context of saturation. We implemented them in Zipperposition~2, which took part
in CASC-J10 \cite{gs-21-cascj10} in 2020. This time, our prover won the division, proving 84\% of
the problems, a whole 20~percentage points ahead of the runner-up, Satallax
3.4.

In this article, we describe the main techniques that explain this reversal
of fortunes. They cover most parts of a modern higher-order theorem prover, from
preprocessing to additional calculus rules to heuristics to backend
integration. We use a newer version of Zipperposition, based on a newer
calculus:\ Instead of \lsup{} augmented with ad hoc Boolean rules,
we work with \emph{\osup} \cite{bbtv-21-full-ho-sup}, a principled extension of superposition to full
higher-order logic, including an interpreted Boolean type.

Many higher-order problems extensively use symbol definitions to simplify
their representation. We describe several ways to exploit the definitions,
%of which the most successful is
such as turning them into rewrite rules (Sect.~\ref{sec:satfol:ho-tech:preprocessing}).
%Interesting patterns can be observed in various higher-order problem encodings.
%We show how we can exploit these to simplify problems (Sect.~\ref{sec:satfol:ho-tech:preprocessing}).
%
By working on formulas rather than clauses, tableau techniques take a more
holistic view of a higher-order problem.
Through its support for delayed clausification and, more generally,
calculus-level formula manipulation, \osup{} enables us to
simulate most successful tableau techniques in a saturating prover
(Sect.~\ref{sec:satfol:ho-tech:formulas}). This calculus also supports \emph{Boolean selection
functions}, a mechanism that allows us to choose on which Boolean subterms
to perform inferences first.
We implemented some Boolean selection functions and
evaluated them (Sect.~\ref{sec:satfol:ho-tech:bool-select}).

The main drawback of both \lsup{} variants compared with combinatory
superposition is that they rely on rules that enumerate possibly infinite sets
of unifiers. We describe a mechanism that interleaves infinitely
branching inferences with the standard saturation process
(Sect.~\ref{sec:satfol:ho-tech:infinite-branching}). The prover
retains the same behavior as
before on first-order problems, smoothly scaling with
increasing numbers of higher-order clauses.
%
We also propose some heuristics to curb the explosion induced by highly
prolific calculus rules (Sect.~\ref{sec:satfol:ho-tech:explosiveness}).

Using first-order backends to finish the proof is common practice in
higher-order reasoning. Since \lsup{} coincides with standard
superposition on first-order clauses, invoking backends may
seem redundant; yet Zipperposition is nowhere as efficient as E
\cite{scv-19-e23} or Vampire \cite{lkav-13-vampire}, so invoking a more
efficient backend does make sense. We describe how to achieve a balance
between allowing native higher-order reasoning and
delegating reasoning to a backend (Sect.~\ref{sec:satfol:ho-tech:backends}).
%
Finally, we compare Zipperposition~2 with other provers on all monomorphic
higher-order TPTP benchmarks \cite{gs-17-tptp} to perform a more extensive
evaluation than at CASC (Sect.~\ref{sec:satfol:ho-tech:comparison}). Our evaluation
corroborates the competition results.

This article is an extended version of a paper accepted at CADE-28
\cite{making-ho-work}. Compared with the conference paper, it
describes a new preprocessing technique, explores the effects of Boolean
selection functions, evaluates %some previously unevaluated
more techniques,
introduces new benchmark sets, and presents more examples.

>>>>

%% TACAS: 16 pp excluding bibliography

\pagestyle{plain}
\PassOptionsToPackage{hyphens}{url}

\usepackage[T1]{fontenc}
\usepackage[misc]{ifsym}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{booktabs}
\usepackage{cite}
\usepackage{color}
\usepackage{enumerate}
\usepackage{enumitem}
%\usepackage{stmaryrd}
%\usepackage{textcomp}
\usepackage{url}
%\usepackage{xparse}
%\usepackage{xpatch}
%\usepackage{mathtools}
\usepackage{microtype}
\usepackage{array}
%\usepackage{nicefrac}
%\usepackage{subcaption}
%\usepackage{tabu}
\usepackage{prftree}
%\usepackage{float}
%\usepackage{bm}
%\usepackage{graphicx}
%\usepackage{wrapfig}
\usepackage{multirow}
%\usepackage{soul}
\usepackage{fancyvrb}
\usepackage{siunitx}
%\let\vec=\relax
%\usepackage{fdsymbol}
%\let\longleadsto=\leadsto
\usepackage{pifont}
\usepackage{hyphenat}
\usepackage{algorithmicx}
\usepackage{regexpatch}
\usepackage{algorithm}% http://ctan.org/pkg/algorithms
\usepackage[noend]{algpseudocode}% http://ctan.org/pkg/algorithmicx
\xpatchcmd{\algorithmic}{\labelsep 0.5em}{\labelsep 3em}{\typeout{Success!}}{\typeout{Oh dear!}}

\hyphenation{ab-strac-tion}

%% for algorithms
%\usepackage[plain]{algorithm}
%\usepackage[noend]{algpseudocode}

\usepackage[
   a4paper,
   pdftex,
%   pdftitle={Superposition for Lambda-Free Higher-Order Logic},
%   pdfauthor={Alexander Bentkamp, Jasmin Blanchette, Simon Cruanes, and Uwe Waldmann},
   pdfkeywords={},
   pdfborder={0 0 0},
   draft=false,
   bookmarksnumbered,
   bookmarks,
   bookmarksdepth=2,
   bookmarksopenlevel=2,
   bookmarksopen]{hyperref}

\usepackage{tikz}
\usepackage{mdframed}
%\usepackage{tikz-qtree}
%\usetikzlibrary{arrows,shapes,positioning,shadows,trees,calc}

%%% BEGIN Times font
%\usepackage{mathptmx}
%\usepackage[scaled=.82]{beramono}
%\usepackage[scaled=.865]{helvet}
%%\usepackage{avant}
%\DeclareSymbolFont{letters}{OML}{txmi}{m}{it}
%%% END Times font

\makeatletter
\g@addto@macro{\UrlBreaks}{\UrlOrds\do\=\do\_}
\makeatother

\hyphenation{counter-example counter-examples Isa-belle Ma-try-osh-ka
  frame-work frame-works non-empty sko-lem-iza-tion super-posi-tion
  over-approx-imate over-approx-imated auto-schedule}

@@

\let\labelitemi=\labelitemii %% CHEAT!

%%% BEGIN poor (wo)man's algorithms
% \newcommand\Function[2]{\textbf{function} \textsc{#1}(#2) \textbf{is} \\}
% \newcommand\Procedure[2]{\textbf{procedure} \textsc{#1}(#2) \textbf{is} \\}
% \newcommand\If[1]{\textbf{if} #1 \textbf{then} \\}
% \newcommand\IfWoThen[1]{\textbf{if} #1 \\}
% \newcommand\Then[1]{\phantom{\textbf{if}} \quad #1 \textbf{then} \\}
% \newcommand\Else{\textbf{else} \\}
% \newcommand\ElsIf[1]{\textbf{else} \textbf{if} #1 \textbf{then} \\}
% \newcommand\ForTo[2]{\textbf{for} #1 \textbf{to} #2 \textbf{do} \\}
% \newcommand\Do{\textbf{do} \\}
% \newcommand\Forever{\textbf{forever} \textbf{do} \\}
% \newcommand\WhileOfDo[1]{\textbf{while} #1}
% \newcommand\ForDownto[2]{\textbf{for} #1 \textbf{downto} #2 \textbf{do} \\}
% \newcommand\While[1]{\textbf{while} #1 \textbf{do} \\}
% \newcommand\Return{\textbf{return} }

% \newcommand\q{\noindent\hbox{}\quad}
% \newcommand\qq{\q\q}
% \newcommand\qqq{\qq\q}
% \newcommand\qqqq{\qqq\q}
% \newcommand\qqqqq{\qqqq\q}
% \newcommand\qqqqqq{\qqqqq\q}
% \newcommand\qqqqqqq{\qqqqqq\q}
% \newcommand\qqqqqqqq{\qqqqqqq\q}
% \newcommand\qqqqqqqqq{\qqqqqqqq\q}

% \newlength\quotexwidth
% \quotexwidth=\textwidth
% \addtolength\quotexwidth{-\leftmargin}

\newcommand\NumberOK[1]{#1}
\newcommand\NumberNOK[1]{\colorbox{red}{#1}}

\newcommand\ourpara[1]{\paragraph{\upshape\bfseries#1.}}

\newenvironment{quotex}
 {\list{}{\rightmargin0pt}%
  \item\relax}
 {\endlist}
%%% END poor (wo)man's algorithms

\newcommand{\cmark}{\text{\ding{51}}}
\newcommand{\xmark}{\text{\ding{55}}}

\newcommand\CRS{\confrep{\texttt{ConjectureRelativeSymbol}}{\texttt{CRS}}}

\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

\newcommand\HLINE{\\[.25\jot]\hline\noalign{\vskip.5\jot}}
\newcommand\infname[1]{\textsc{#1}}
\newcommand{\infer}[3]{\prftree[r]{{\small\infname{#1}}}{\strut#2}{\strut#3}}
\newcommand{\inferdouble}[3]{\prftree[d,r]{{\small\infname{#1}}}{\strut#2}{\strut#3}}

\newcommand{\hooklongrightarrow}{\lhook\joinrel\longrightarrow}

\newcommand\Unifarrow{\Longrightarrow}
\newcommand\Matcharrow{\Longrightarrow}
\newcommand\Pdtarrow{\leadsto}
\newcommand\Matchiiarrow{\hooklongrightarrow}

% Most general unifier
\DeclareMathOperator{\mgu}{mgu}

\def\negvthinspace{\kern-0.083333em}
\def\vthinspace{\kern+0.083333em}
\def\vvthinspace{\kern+0.0416667em}
\def\negvvthinspace{\kern-0.0416667em}
%\def\hypsep{\hskip0.75em}
\def\hypsep{\quad}
\def\rulesep{\qquad}

\spnewtheorem{theoremx}[theorem]{Theorem}{\bfseries}{\slshape}
\spnewtheorem{corollaryx}[theorem]{Corollary}{\bfseries}{\slshape}
\spnewtheorem{lemmax}[theorem]{Lemma}{\bfseries}{\slshape}

\newcommand\ehohii{$\lambda$E}

\newcommand{\eq}{\approx}
\newcommand{\noteq}{\not\eq}

\newcommand\UNIF{\mathrel{\smash{\stackrel{\lower.1ex\hbox{\ensuremath{\scriptscriptstyle ?}}}{=}}}}
%\newcommand\UNIF{=^?}
\newcommand\MATCH{\mathrel{{\scriptstyle\lesssim}^{\scriptscriptstyle{?\!}}}}

%\newcommand\UNIF{=^{?\!}}
%\newcommand\MATCH{\lesssim^{?\!}}
%\newcommand\MATCH{\mathrel{\smash{\stackrel{\lower.1ex\hbox{\ensuremath{\scriptscriptstyle ?}}}{\lesssim}}}}

\newcommand\Var{\mathcal{V}\negvthinspace\mathit{ar}}
\newcommand\Term{\mathcal{T}\kern-.7ex\mathit{erm}}
\newcommand\Gfpf{\mathcal{G}\kern-.3ex\mathit{fpf}}
\newcommand\GfpfRTL{\Gfpf'}
\newcommand\Fp{\mathcal{F}\kern-.3ex\mathit{p}}
\newcommand\FpRTL{\Fp'}

%\renewcommand\AA{\textbf{\textsf{A}}}
%\newcommand\BB{\textbf{\textsf{B}}}
%\newcommand\NN{\textbf{\textsf{N}}}

\renewcommand\AA{{\textsf{A}}}
\newcommand\BB{{\textsf{B}}}
\newcommand\NN{{\textsf{N}}}

\definecolor{verylightgray}{gray}{0.9}
\definecolor{pdtgreen}{HTML}{97D077}
\definecolor{pdtorange}{HTML}{D79B00}

\DeclareRobustCommand{\hlorange}[1]{{\sethlcolor{orange}\hl{#1}}}
\DeclareRobustCommand{\hlgreen}[1]{{\sethlcolor{green}\hl{#1}}}
\DeclareRobustCommand{\hlcolor}[2]{{\sethlcolor{#1}\hl{#2}}}

% for Springer theorems
\spnewtheorem{examplex}{Example}{\bfseries}{\rmfamily}
\spnewtheorem{definitionx}[theorem]{Definition}{\bfseries}{\rmfamily}

% probably can be set up differently
%\theoremstyle{definition}
%\newtheorem{definitionx}{Definition}

%\theoremstyle{definition}
%\newtheorem{examplex}{Example}

%-- from my thesis
%\theoremstyle{definition}
%\newtheorem{remark}{Remark}

\newcommand\afterDot{\,} %%% the Springer style is a bit broken---this makes things a bit more tolerable
\def\cpp{C\nobreak\raisebox{.1ex}{+}\nobreak\raisebox{.1ex}{+}}
\newcommand\appE{\ensuremath{\cst{@}{+}\text{E}}}

\newcommand\cst[1]{\mathsf{#1}}

%%%%%%%%%%%%%%%%%%%%%%
% conf and rep
\newcommand{\confrep}[2]{\begin{conf}#1\end{conf}\begin{rep}#2\end{rep}}
\newcommand{\lfhol}[0]{$\lambda$fHOL}
\newcommand{\appvar}[0]{\ensuremath{@}}

% stolen from FOBOOLSUP paper
\newcommand{\ifalse}{\pmb\bot}
\newcommand{\itrue}{\pmb\top}
\newcommand{\inot}{\pmb{\neg\,}}
\newcommand{\iand}{\pmb\land}
\newcommand{\ior}{\pmb\lor}
\newcommand{\iimplies}{\pmb\rightarrow}
\newcommand{\iequiv}{\pmb\leftrightarrow}
\newcommand{\iforall}{\pmb\forall}
\newcommand{\iexists}{\pmb\exists}
\newcommand{\ieq}{\pmb\approx}
\newcommand{\ineq}{\pmb{\not\approx}}

\newcommand{\comb}[1]{\ensuremath{\mathsf{#1}}}
\newcommand{\neglit}[1]{\ensuremath{#1 \not\eq \itrue}}
\newcommand{\poslit}[1]{\ensuremath{#1 \eq \itrue}}
\newcommand{\lland}{\mathrel\land}
\newcommand{\llor}{\mathrel\lor}
\newcommand{\ccup}{\mathrel\cup}
\newcommand{\ccap}{\mathrel\cap}

\newcommand{\db}[1]{\ensuremath{\mathbf{#1}}}
\newcommand{\dbvar}[1]{\ensuremath{\textbf{\textit{#1}}}}
\newcommand{\internallam}{\ensuremath{\texttt{LAM}}}
\newcommand{\internalat}{\ensuremath{\texttt{@}}}

\newcommand{\lambdabf}{\pmb{\lambda}} %%% TYPESETTING
\newcommand{\betabf}{\pmb{\beta}} %%% TYPESETTING
\newcommand{\etabf}{\pmb{\eta}} %%% TYPESETTING

@@

\newcommand\CORR{\textsuperscript{(\vthinspace\Letter\vthinspace)}}

\confrep
{\title{Extending a High-Performance Prover to Higher-Order Logic}}
{\title{Extending a High-Performance Prover to Higher-Order Logic \\ (Technical Report)}}
%{\title{Extending a Brainiac Prover to \\ Full Higher-Order Logic}}
%{\title{Extending a Brainiac Prover to \\ Full Higher-Order Logic\\(Technical Report)}}

\author{Petar Vukmirovi\'c\inst{1}\CORR
  \and
  Jasmin Blanchette\inst{1,2}
  \and
  Stephan Schulz\inst{3}
  }

\institute{
  Vrije Universiteit Amsterdam, Amsterdam, the Netherlands \\
%  \email{p.vukmirovic@vu.nl}
   \email{\{p.vukmirovic,j.c.blanchette\}@vu.nl}
%   \and Max-Planck-Institut f\"ur Informatik, Saarland Informatics Campus, Saarbr\"ucken,~Germany \\
%   \email{jasmin.blanchette@mpi-inf.mpg.de}
  \and Universit\'e de Lorraine, CNRS, Inria, LORIA, Nancy, France \\
    \email{jasmin.blanchette@loria.fr}
  \and DHBW Stuttgart, Stuttgart, Germany \\
    \email{stephan.schulz@dhbw-stuttgart.de}
}

@@

\begin{abstract}%
  The automatic discharge of tedious subgoals is high on the wishlist of many
  users of proof assistants. Some proof assistants discharge such goals
  by translating them to first-order logic and invoking an efficient prover on
  them, but much is lost in translation. As an alternative,
  we propose to extend first-order provers with native support for
  higher-order features. Building on our extension of E to $\lambda$-free
  higher-order logic, we now extend E to full higher-order logic.
  The resulting prover is the second strongest on both TPTP benchmarks and
  benchmarks coming from a proof assistant.
  %It also incurs no overhead on first-order problems.  -- sounds like a detail --JB
\end{abstract}

\setcounter{footnote}{0}

\section{Introduction}
\label{sec:introduction}

In the last few decades, proof assistants have become indispensible tools for
developing trustworthy formal proofs. They are used both in academia to verify
mathematical theories \cite{th-05-flyspeck} and in
industry to verify the correctness of hardware \cite{ckmg-99-hw-verification}
and software \cite{xl-09-compcert, gk-09-sel4, rgea-16-certikos}.
%
However, due to the lack of strong built-in proof automation, proving
seemingly simple goals can be a tedious manual task. To mitigate
this, many proof assistants include a subsystem such as CoqHammer,
HOL(y)Hammer, or Sledgehammer \cite{bkpu-16-hammering} that translates
higher-order goals to first-order logic and passes them to efficient
first-order automatic provers. If a first-order prover succeeds, the proof is
reconstructed and the goal is closed.

Unfortunately, the translation of higher-order constructs is clumsy and leads
to poor performance on goals that require higher-order reasoning.
Using native higher-order provers
such as %Leo-III \cite{sb-21-leo3} or   --- the reference is about Leo-II and Satallax --JB
Satallax \cite{cb-12-satallax} as backends
is not necessarily a solution because they are much less optimized than their
first-order counterparts \cite{ns-13-leo2sh}. To bridge this gap, in 2016
we proposed to develop a new generation of higher-order provers that
%are not built from the ground up, but rather
extend the arguably most successful first-order calculus, superposition, to
higher-order logic, starting from a position of strength.

Our research has focused on three milestones:\ supporting $\lambda$-free
higher-order logic, adding support for $\lambda$-terms, and adding support for
first-class Boolean terms. In 2019, we extended the state-of-the-art first-order
prover E \cite{scv-19-e23} with a $\lambda$-free superposition calculus
\cite{section-ehoh}, obtaining a version of E called Ehoh, as a stepping stone
towards full higher-order logic. Together with Bentkamp, Tourret, and Waldmann,
we have since developed calculi, called \emph{$\lambda$-superposition}, corresponding
to the other two milestones \cite{bbtvw-21-sup-lam,bbtv-21-full-ho-sup} and
implemented them in the experimental superposition prover Zipperposition
\cite{sc-15-simon-phd}. This prover is implemented in OCaml and not nearly as
efficient as E. Nevertheless, it has won the higher-order division of the CASC
theorem prover competition \cite{gs-21-cascj10} in 2020 and 2021 by a wide
margin, ending nearly a decade of Satallax domination.

In this \paper, we fulfill a three-year-old promise: We present the extension of
Ehoh to full higher-order logic (Sect.~\ref{sec:logic}) using incomplete variants
of $\lambda$-superposition. We call this prover \ehohii.
%
The $\lambda$-superposition calculi were
previously implemented in Zipperposition, and
extensive experiments with various heuristic choices have been performed
\cite{section-making-ho-work}. In \ehohii{}'s implementation, we used
these experiences to choose a set of effective rules that could easily be
retrofitted into an originally first-order prover. Another principle that guided 
the design of \ehohii{} was \emph{gracefulness}: we made sure that our changes
do not impact already strong first-order performance of Ehoh.

An simpler approach to support higher-order logic is to implement
combinatory superposition, which is based on \textsf{SKBCI}-combinators
and implemented in Vampire \cite{br-20-full-sup-w-combs}.
These can be supported as a thin layer on top of $\lambda$-free terms. However,
in unpublished experiments, we found that $\lambda$-superposition performs
substantially better, corroborating the CASC results, so we
decided to make a more profound change to Ehoh and implement
$\lambda$-superposition.

% We
% also used the experience of fine-tuning the calculi
% \cite{section-making-ho-work} in Zipperposition to carefully choose which
% calculus extensions and heuristics to use in E.

One of the main challenges we faced was retrofitting $\lambda$-terms in Ehoh's
term representation (Sect.~\ref{sec:terms}). Furthermore, Ehoh's main inference
engine was designed with the assumption that it will be used with
inferences that compute a most general unifier. We
implementd a higher-order unification procedure \cite{unif-section}
that can return multiple unifiers (Sect.~\ref{sec:unif-match-index}) and
integrated it in the inference engine. Finally, we extended and adapted the
superposition rule, resulting in an incomplete, pragmatic variant of
$\lambda$-superposition (Sect.~\ref{sec:calculus}).

% most likely text
We \NumberOK{evaluated} \ehohii{} on a selection of proof assistants benchmarks
as well as all higher-order theorems in the TPTP library \cite{gs-17-tptp}
(Sect.~\ref{sec:eval}). We found
that \ehohii{} clearly outperforms Ehoh on all benchmarks. Compared
with other higher-order provers, it is second
only to the cooperative version of Zipperposition, which employs Ehoh as a
backend. A fairer comparison without the backend puts \ehohii{} in the
first place for both benchmark suites.
We also compared the performance of \ehohii{} with E on first-order
%%% @PETAR: I made this slightly sentence shorter and more forceful. Please
%%% check if you agree. --JB
problems and found that little overhead has been introduced by the
extension to higher-order logic.

\kern-1pt %% TYPESETTING hack

\section{Logic} %---0.5p
\label{sec:logic}

% TO BE CHANGED AS NECESSARY
Our target logic is monomorphic classical higher-order logic with Hilbert choice.
The following text is partly based on Section 2 of Vukmirovi\'c et al.~\cite{section-making-ho-work}.

Terms $s, t, u, v$ are inductively defined as free variables $F, X, \ldots$, bound variables $x,
y, z, \dotsc$, constants $\cst{f}, \cst{g},\allowbreak \cst{a},\allowbreak \cst{b},
\dotsc$, applications $s \, t$, and $\lambda$-abstractions $\lamx{s}$.
The syntactic distinction between free and bound variables
gives rise to \emph{loose bound variables} (e.g., $y$ in $\lamx{y \, \cst{a}}$)
\cite{tn-93-patterns}.
%We shorten iterated application $s \, t_1 \, \cdots \, t_n$ to $s \,
%\overline{t}_n$ and $\lambda$-abstraction $\lambda x_1. \, \cdots \, \lambda
%x_n. \> s$ to $\lam{\overline{x}_n}{s}$.

We let $s \, \overline{t}_n$ stand for $s \, t_1 \, \ldots \, t_n$ and
$\lam{\overline{x}_n}{s}$ for $\lambda x_1. \ldots \lambda x_n. \> s$.
Every $\beta$-normal term can be written as $\lam{\overline{x}_m}{s \,
\overline{t}_n}$, where $s$ is not an application; we call $s$ the \emph{head}
of the term. If $s$ is a free variable, we call the term \emph{flex};
otherwise, the term is \emph{rigid}. If the type of a term is of the form $\tau_1 \to \cdots \to
\tau_n \to o$, where $o$ is the distinguished Boolean type and $n \ge 0$, we
call it a \emph{predicate}. A literal $l$ is an equation $s \eq t$ or
a disequation $s \not\eq t$. A clause is a finite multiset of literals,
interpreted and written disjunctively:\ $l_1 \llor \cdots \llor l_n$.
Logical symbols that may occur within terms are written in bold:
$\pmb{\neg}, \iand, \ior, \iimplies, \iequiv, \dots$. Predicate
literals are encoded as (dis)equations with $\itrue$ based on their sign; for
example, $\cst{even}(x)$ is encoded as $\poslit{\cst{even}(x)}$, and
$\neg\,\cst{even}(x)$ is encoded as $\neglit{\cst{even}(x)}$.

\section{Terms}
\label{sec:terms}

Ehoh's term representation is centered around perfect term sharing %or exa
\cite{ls-01-shared}: Any two structurally identical terms are
guaranteed to be the same object in memory. This is achieved through term
\emph{cells}, which represent individual terms. Each cell has (among other fields)
(1)~\texttt{f\_code}, an integer corresponding to the symbol at the head of the term (negative
if the head is a free variable, positive otherwise); (2)~\texttt{num\_args},
corresponding to the number of arguments applied to the head; and (3)~\texttt{args},
a size-\texttt{num\_args} array of pointers to argument terms. We
use the first-order notation $\cst{f}(s_1, \ldots, s_n)$ to denote a cell whose
\texttt{f\_code} corresponds to $\cst{f}$, \texttt{num\_args} equals $n$, and
\texttt{args} points to the cells for $s_1, \ldots s_n$.
